Routines and Regret: An Examination of Behavioral Norms and Emotional Responses

### 2024-02-15 paper/paper.qmd
modify
style

# JAMOVI requires factors, while R imports as numeric. So, need to convert from numeric to factor.
data$Sc1_regret<- factor(data$Sc1_regret)
data$sc1_socnorms1<- factor(data$sc1_socnorms1)
data$sc1_socnorms2<- factor(data$sc1_socnorms2)
data$sc1_combinednorms<- factor(data$sc1_combinednorms)

# Let's label the values better, so it's easier to understand the output.
data$Sc1_regret <- ordered(data$Sc1_regret, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms1 <- ordered(data$sc1_socnorms1, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms2 <- ordered(data$sc1_socnorms2, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_combinednorms <- ordered(data$sc1_combinednorms, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))

# R's way of doing binomial
# we need to count, while excluding NAs
x1 <- sum((data$Sc1_regret[!is.na(data$Sc1_regret)])=="Exception Jones")
n1 <- length(data$Sc1_regret[!is.na(data$Sc1_regret)])
prop.test(x1, n1, p=0.5, correct = FALSE)
result <- prop.test(x1, n1, p = 0.5, correct = FALSE)
p1 <- result$estimate["p"]

x2 <- sum((data$sc1_socnorms1[!is.na(data$sc1_socnorms1)])=="Exception Jones")
n2 <- length(data$sc1_socnorms1[!is.na(data$sc1_socnorms1)])
prop.test(x2, n2, p=0.5, correct = FALSE)
result <- prop.test(x2, n2, p = 0.5, correct = FALSE)
p2 <- result$estimate["p"]

x3 <- sum((data$sc1_socnorms2[!is.na(data$sc1_socnorms2)])=="Exception Jones")
n3 <- length(data$sc1_socnorms2[!is.na(data$sc1_socnorms2)])
prop.test(x3, n3, p=0.5, correct = FALSE)
result <- prop.test(x3, n3, p = 0.5, correct = FALSE)
p3 <- result$estimate["p"]

x4 <- sum((data$sc1_combinednorms[!is.na(data$sc1_combinednorms)])=="Exception Jones")
n4 <- length(data$sc1_combinednorms[!is.na(data$sc1_combinednorms)])
prop.test(x4, n4, p=0.5, correct = FALSE)
result <- prop.test(x4, n4, p = 0.5, correct = FALSE)
p4 <- result$estimate["p"]

# create a data frame to save the data
table_data <- data.frame(
  ` ` = c("Routine Smith", "Exception Jones"),
  Regret_Counts = c(n1 - x1, x1),
  Regret_Percent = c(round(100*(1 - p1), digits = 1), round(100*p1, digits = 1)),
  Social_Norm_Injunctive_Counts = c(n2 - x2, x2),
  Social_Norm_Injunctive_Percent = c(round(100*(1 - p2), digits = 1), round(100*p2, digits = 1)),
  Social_Norm_Descriptive_Counts = c(n3 - x3, x3),
  Social_Norm_Descriptive_Percent = c(round(100*(1 - p3), digits = 1), round(100*p3, digits = 1)),
  Negative_Affect_Counts = c(n4 - x4, x4),
  Negative_Affect_Percent = c(round(100*(1 - p4), digits = 1), round(100*p4, digits = 1))
)
```

```{r}
#| echo: false
#| label: tbl-frequency_and_proportion_exp1
#| tbl-cap: "Frequency and proportion of regret, social norms, and negative affect for exp1."
# create the table with kable()
kable(table_data, format = "latex", col.names = c(" ", "Regret", "%", "Soc_norm1", "%", "Soc_norm2", "%", "Negetive affect", "%"), align = 'c')
```

@@

```{r}
#| echo: false
#| label: fig-proportion_exp1
#| fig-cap: "Proportion of regret, social norms, and negative affect for exp1."
# Data in percentage for Smith and Jones we gained from the table we created
# Ensure that the percentages are divided by 100 to be in the correct format
data2 <- data.frame(
  Category = c('Regret', 'Social norms (injunctive)', 'Social norms (descriptive)', 'Negative affect'),
  Smith = c(7.9 / 100, 4.7 / 100, 90.6 / 100, 7.3 / 100), # Convert to proportions
  Jones = c(92.1 / 100, 95.3 / 100, 9.4 / 100, 92.7 / 100) # Convert to proportions
)

# Melt the data for use with ggplot2
data_melted <- melt(data2, id.vars='Category')

# Create the bar plot with specified colors and corrected percentage values
ggplot(data_melted, aes(x = Category, y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c('Smith' = '#1f77b4', 'Jones' = '#d62728')) +
  theme_minimal() +
  labs(x = '', y = 'Percentage (%)', fill = '') +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = 'bottom',
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_y_continuous(labels = percent_format()) + # This will ensure y-axis is formatted as percentage
  ggtitle('Proportions for regret, social norms, and negative affect')

```

@@

# Create a table
kable(final_data, col.names = c("", "Choice", "Count", "%"), align = c('l', 'l', 'r', 'r'))
```

@@

# Data in percentage for Adams and White we gained from the table we created
# Ensure that the percentages are divided by 100 to be in the correct format
data <- data.frame(
  Category = c('Regret', 'Luck'),
  Adams_Routine = c(19.0 / 100, 33.3 / 100), # Convert to proportions
  White_Exception = c(81.0 / 100, 66.7 / 100) # Convert to proportions
)

# Melt the data for use with ggplot2
data_melted <- melt(data, id.vars='Category')

# Create the bar plot with specified colors and corrected percentage values
ggplot(data_melted, aes(x = Category, y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c('Adams_Routine' = '#1f77b4', 'White_Exception' = '#d62728')) +
  theme_minimal() +
  labs(x = '', y = 'Percentage (%)', fill = '') +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = 'bottom',
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_y_continuous(labels = percent_format()) + # This will ensure y-axis is formatted as percentage
  ggtitle('Proportions for regret and luck')

```

@@

head(simulated_exp2) |>
 kable(
    col.names = c("id", "regret", "luck"),
    booktabs = TRUE
  )

```

@@

#### Load data and perform statistical analysis####
# setting formatting options
options(scipen=999.99, digits =7)

@@

#Check the top six rows of data of interest
dataset_selected <- data %>% select(Sc1_regret, sc1_socnorms1, sc1_socnorms2, sc1_combinednorms, Sc2_regret, Sc2_lucky) 
kable(head(dataset_selected))
```

@@

#Show demographics
#Age
data$gender
table(data$gender)
#Gender
data$age
data$age[data$age==99] <- NA
mean(data$age, na.rm = TRUE)
sd(data$age, na.rm = TRUE)
```

@@

# JAMOVI requires factors, while R imports as numeric. So, need to convert from numeric to factor.
data$Sc1_regret<- factor(data$Sc1_regret)
data$sc1_socnorms1<- factor(data$sc1_socnorms1)
data$sc1_socnorms2<- factor(data$sc1_socnorms2)
data$sc1_combinednorms<- factor(data$sc1_combinednorms)

# Let's label the values better, so it's easier to understand the output.
data$Sc1_regret <- ordered(data$Sc1_regret, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms1 <- ordered(data$sc1_socnorms1, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms2 <- ordered(data$sc1_socnorms2, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_combinednorms <- ordered(data$sc1_combinednorms, levels = c(1,2), labels = c("Exception Jones", "Routine Smith"))

@@

# Let's run the JAMOVI imported syntax 
# Descriptives for the main variables.
# Plots appear in the R Studio Plots section
jmv::descriptives(
  data=data,
  vars=c(
    "Sc1_regret",
    "sc1_socnorms1",
    "sc1_socnorms2",
    "sc1_combinednorms"),
  freq=TRUE)

```

@@

# binomial Z
jmv::propTest2(
  data=data,
  vars=c(
    "Sc1_regret",
    "sc1_socnorms1",
    "sc1_socnorms2",
    "sc1_combinednorms"),
  ci=TRUE)

```

@@

#Measure correction: All particpants who indicated (5- somewhat agree) on the question regarding random chance (data$Sc2_random_1 or data$Sc2_random_2) had accidently assigned the value '56' instead of '5' in Qualtrics
data$Sc2_random_1[data$Sc2_random_1==56] <- 5
data$Sc2_random_2[data$Sc2_random_2==56] <- 5

# Conversion from numeric to factors 
data$Sc2_regret<- factor(data$Sc2_regret)
data$Sc2_lucky<- factor(data$Sc2_lucky)

# Label values
data$Sc2_regret <- ordered(data$Sc2_regret, levels = c(1,2), labels = c("Routine Adams", "Exception White"))
data$Sc2_lucky <- ordered(data$Sc2_lucky, levels = c(1,2), labels = c("Adams less lucky", "White less lucky"))
names (data$Sc2_random_1) <- c("Strongly disagree", "Disagree", "Somewhat disagree", "Neither agree nor disagree", "Somewhat agree", "Agree", "Strongly agree")
names (data$Sc2_random_2) <- c("Strongly disagree", "Disagree", "Somewhat disagree", "Neither agree nor disagree", "Somewhat agree", "Agree", "Strongly agree")

@@

# Descriptives for main variables
jmv::descriptives(
  data=data,
  vars=c(
    "Sc2_regret",
    "Sc2_random_1",
    "Sc2_random_2",
    "Sc2_lucky"),
  freq=TRUE)
```

```{r}
#| echo: false
#| warning: false
#| include: false
# binomial Z
jmv::propTest2(
  data=data,
  vars=c(
    "Sc2_regret",
    "Sc2_lucky"),
  ci=TRUE)
```

@@

# Modify the check to correctly account for NA values and check whether the 
# values are one of the "Routine Smith", "Exception Jones".
all(data$Sc1_regret %in% allowed_values_exp1 | is.na(data$Sc1_regret))== TRUE
all(data$sc1_socnorms1 %in% allowed_values_exp1 | is.na(data$sc1_socnorms1))== TRUE
all(data$sc1_socnorms2 %in% allowed_values_exp1 | is.na(data$sc1_socnorms2))== TRUE
all(data$sc1_combinednorms %in% allowed_values_exp1 | is.na(data$sc1_combinednorms))== TRUE

# EXPERIMENT 2 (Car Accident-Scenario)
# Adjust the allowed_values to not include "NA" as a string
allowed_values_exp2_regret <- c("Routine Adams", "Exception White")
allowed_values_exp2_lucky <- c("Adams less lucky", "White less lucky")
# Modify the check to correctly account for NA values and check whether the 
# values are one of the "Routine Adams", "Exception White".
all(data$Sc2_regret %in% allowed_values_exp2_regret | is.na(data$Sc2_regret))== TRUE
all(data$Sc2_lucky %in% allowed_values_exp2_lucky | is.na(data$Sc2_lucky))== TRUE
```

>>>>

# JAMOVI requires factors, while R imports as numeric. So, need to convert from numeric to factor.
data$Sc1_regret <- factor(data$Sc1_regret)
data$sc1_socnorms1 <- factor(data$sc1_socnorms1)
data$sc1_socnorms2 <- factor(data$sc1_socnorms2)
data$sc1_combinednorms <- factor(data$sc1_combinednorms)

# Let's label the values better, so it's easier to understand the output.
data$Sc1_regret <- ordered(data$Sc1_regret, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms1 <- ordered(data$sc1_socnorms1, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms2 <- ordered(data$sc1_socnorms2, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_combinednorms <- ordered(data$sc1_combinednorms, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))

# R's way of doing binomial
# we need to count, while excluding NAs
x1 <- sum((data$Sc1_regret[!is.na(data$Sc1_regret)]) == "Exception Jones")
n1 <- length(data$Sc1_regret[!is.na(data$Sc1_regret)])
prop.test(x1, n1, p = 0.5, correct = FALSE)
result <- prop.test(x1, n1, p = 0.5, correct = FALSE)
p1 <- result$estimate["p"]

x2 <- sum((data$sc1_socnorms1[!is.na(data$sc1_socnorms1)]) == "Exception Jones")
n2 <- length(data$sc1_socnorms1[!is.na(data$sc1_socnorms1)])
prop.test(x2, n2, p = 0.5, correct = FALSE)
result <- prop.test(x2, n2, p = 0.5, correct = FALSE)
p2 <- result$estimate["p"]

x3 <- sum((data$sc1_socnorms2[!is.na(data$sc1_socnorms2)]) == "Exception Jones")
n3 <- length(data$sc1_socnorms2[!is.na(data$sc1_socnorms2)])
prop.test(x3, n3, p = 0.5, correct = FALSE)
result <- prop.test(x3, n3, p = 0.5, correct = FALSE)
p3 <- result$estimate["p"]

x4 <- sum((data$sc1_combinednorms[!is.na(data$sc1_combinednorms)]) == "Exception Jones")
n4 <- length(data$sc1_combinednorms[!is.na(data$sc1_combinednorms)])
prop.test(x4, n4, p = 0.5, correct = FALSE)
result <- prop.test(x4, n4, p = 0.5, correct = FALSE)
p4 <- result$estimate["p"]

# create a data frame to save the data
table_data <- data.frame(
  ` ` = c("Routine Smith", "Exception Jones"),
  Regret_Counts = c(n1 - x1, x1),
  Regret_Percent = c(round(100 * (1 - p1), digits = 1), round(100 * p1, digits = 1)),
  Social_Norm_Injunctive_Counts = c(n2 - x2, x2),
  Social_Norm_Injunctive_Percent = c(round(100 * (1 - p2), digits = 1), round(100 * p2, digits = 1)),
  Social_Norm_Descriptive_Counts = c(n3 - x3, x3),
  Social_Norm_Descriptive_Percent = c(round(100 * (1 - p3), digits = 1), round(100 * p3, digits = 1)),
  Negative_Affect_Counts = c(n4 - x4, x4),
  Negative_Affect_Percent = c(round(100 * (1 - p4), digits = 1), round(100 * p4, digits = 1))
)
```

```{r}
#| echo: false
#| label: tbl-frequency_and_proportion_exp1
#| tbl-cap: "Frequency and proportion of regret, social norms, and negative affect for exp1."
# create the table with kable()
kable(table_data, format = "latex", col.names = c(" ", "Regret", "%", "Soc_norm1", "%", "Soc_norm2", "%", "Negetive affect", "%"), align = "c")
```

@@

```{r}
#| echo: false
#| label: fig-proportion_exp1
#| fig-cap: "Proportion of regret, social norms, and negative affect for exp1."
# Data in percentage for Smith and Jones we gained from the table we created
# Ensure that the percentages are divided by 100 to be in the correct format
data2 <- data.frame(
  Category = c("Regret", "Social norms (injunctive)", "Social norms (descriptive)", "Negative affect"),
  Smith = c(7.9 / 100, 4.7 / 100, 90.6 / 100, 7.3 / 100), # Convert to proportions
  Jones = c(92.1 / 100, 95.3 / 100, 9.4 / 100, 92.7 / 100) # Convert to proportions
)

# Melt the data for use with ggplot2
data_melted <- melt(data2, id.vars = "Category")

# Create the bar plot with specified colors and corrected percentage values
ggplot(data_melted, aes(x = Category, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("Smith" = "#1f77b4", "Jones" = "#d62728")) +
  theme_minimal() +
  labs(x = "", y = "Percentage (%)", fill = "") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_y_continuous(labels = percent_format()) + # This will ensure y-axis is formatted as percentage
  ggtitle("Proportions for regret, social norms, and negative affect")
```

@@

# Create a table
kable(final_data, col.names = c("", "Choice", "Count", "%"), align = c("l", "l", "r", "r"))
```

@@

# Data in percentage for Adams and White we gained from the table we created
# Ensure that the percentages are divided by 100 to be in the correct format
data <- data.frame(
  Category = c("Regret", "Luck"),
  Adams_Routine = c(19.0 / 100, 33.3 / 100), # Convert to proportions
  White_Exception = c(81.0 / 100, 66.7 / 100) # Convert to proportions
)

# Melt the data for use with ggplot2
data_melted <- melt(data, id.vars = "Category")

# Create the bar plot with specified colors and corrected percentage values
ggplot(data_melted, aes(x = Category, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Adams_Routine" = "#1f77b4", "White_Exception" = "#d62728")) +
  theme_minimal() +
  labs(x = "", y = "Percentage (%)", fill = "") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_y_continuous(labels = percent_format()) + # This will ensure y-axis is formatted as percentage
  ggtitle("Proportions for regret and luck")
```

@@

head(simulated_exp2) |>
  kable(
    col.names = c("id", "regret", "luck"),
    booktabs = TRUE
  )
```

@@

#### Load data and perform statistical analysis####
# setting formatting options
options(scipen = 999.99, digits = 7)

@@

# Check the top six rows of data of interest
dataset_selected <- data %>% select(Sc1_regret, sc1_socnorms1, sc1_socnorms2, sc1_combinednorms, Sc2_regret, Sc2_lucky)
kable(head(dataset_selected))
```

@@

# Show demographics
# Age
data$gender
table(data$gender)
# Gender
data$age
data$age[data$age == 99] <- NA
mean(data$age, na.rm = TRUE)
sd(data$age, na.rm = TRUE)
```

@@

# JAMOVI requires factors, while R imports as numeric. So, need to convert from numeric to factor.
data$Sc1_regret <- factor(data$Sc1_regret)
data$sc1_socnorms1 <- factor(data$sc1_socnorms1)
data$sc1_socnorms2 <- factor(data$sc1_socnorms2)
data$sc1_combinednorms <- factor(data$sc1_combinednorms)

# Let's label the values better, so it's easier to understand the output.
data$Sc1_regret <- ordered(data$Sc1_regret, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms1 <- ordered(data$sc1_socnorms1, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_socnorms2 <- ordered(data$sc1_socnorms2, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))
data$sc1_combinednorms <- ordered(data$sc1_combinednorms, levels = c(1, 2), labels = c("Exception Jones", "Routine Smith"))

@@

# Let's run the JAMOVI imported syntax
# Descriptives for the main variables.
# Plots appear in the R Studio Plots section
jmv::descriptives(
  data = data,
  vars = c(
    "Sc1_regret",
    "sc1_socnorms1",
    "sc1_socnorms2",
    "sc1_combinednorms"
  ),
  freq = TRUE
)
```

@@

# binomial Z
jmv::propTest2(
  data = data,
  vars = c(
    "Sc1_regret",
    "sc1_socnorms1",
    "sc1_socnorms2",
    "sc1_combinednorms"
  ),
  ci = TRUE
)
```

@@

# Measure correction: All particpants who indicated (5- somewhat agree) on the question regarding random chance (data$Sc2_random_1 or data$Sc2_random_2) had accidently assigned the value '56' instead of '5' in Qualtrics
data$Sc2_random_1[data$Sc2_random_1 == 56] <- 5
data$Sc2_random_2[data$Sc2_random_2 == 56] <- 5

# Conversion from numeric to factors
data$Sc2_regret <- factor(data$Sc2_regret)
data$Sc2_lucky <- factor(data$Sc2_lucky)

# Label values
data$Sc2_regret <- ordered(data$Sc2_regret, levels = c(1, 2), labels = c("Routine Adams", "Exception White"))
data$Sc2_lucky <- ordered(data$Sc2_lucky, levels = c(1, 2), labels = c("Adams less lucky", "White less lucky"))
names(data$Sc2_random_1) <- c("Strongly disagree", "Disagree", "Somewhat disagree", "Neither agree nor disagree", "Somewhat agree", "Agree", "Strongly agree")
names(data$Sc2_random_2) <- c("Strongly disagree", "Disagree", "Somewhat disagree", "Neither agree nor disagree", "Somewhat agree", "Agree", "Strongly agree")

@@

# Descriptives for main variables
jmv::descriptives(
  data = data,
  vars = c(
    "Sc2_regret",
    "Sc2_random_1",
    "Sc2_random_2",
    "Sc2_lucky"
  ),
  freq = TRUE
)
```

```{r}
#| echo: false
#| warning: false
#| include: false
# binomial Z
jmv::propTest2(
  data = data,
  vars = c(
    "Sc2_regret",
    "Sc2_lucky"
  ),
  ci = TRUE
)
```

@@

# Modify the check to correctly account for NA values and check whether the
# values are one of the "Routine Smith", "Exception Jones".
all(data$Sc1_regret %in% allowed_values_exp1 | is.na(data$Sc1_regret)) == TRUE
all(data$sc1_socnorms1 %in% allowed_values_exp1 | is.na(data$sc1_socnorms1)) == TRUE
all(data$sc1_socnorms2 %in% allowed_values_exp1 | is.na(data$sc1_socnorms2)) == TRUE
all(data$sc1_combinednorms %in% allowed_values_exp1 | is.na(data$sc1_combinednorms)) == TRUE

# EXPERIMENT 2 (Car Accident-Scenario)
# Adjust the allowed_values to not include "NA" as a string
allowed_values_exp2_regret <- c("Routine Adams", "Exception White")
allowed_values_exp2_lucky <- c("Adams less lucky", "White less lucky")
# Modify the check to correctly account for NA values and check whether the
# values are one of the "Routine Adams", "Exception White".
all(data$Sc2_regret %in% allowed_values_exp2_regret | is.na(data$Sc2_regret)) == TRUE
all(data$Sc2_lucky %in% allowed_values_exp2_lucky | is.na(data$Sc2_lucky)) == TRUE
```
